---
title: "Hands_on_Ex01"
---

## 1.2 Installling and loading the required libraries

## 1.2.1 Importing data

```{r}
library(readr) # Load the readr library
exam_data <- read_csv("data/Exam_data.csv")
```

## 1.3 Introducing ggplot

## 1.3.1 R Graphics VS ggplot

```{r}
hist(exam_data$MATHS)
```

```{r}
library(ggplot2)
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

## 1.4 Grammar of Graphics

```{r}
library(ggplot2)
```

## 1.4.1 A Layered Grammar of Graphics

## 1.5 Essential Grammatical Elements in ggplot2: data

```{r}
ggplot(data=exam_data)
```

## 1.6 Essential Grammatical Elements in ggplot2: Aesthetic mappings

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS))
```

## 1.7 Essential Grammatical Elements in ggplot2: geom

## 1.7.1 Geometric Objects: geom_bar

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

## 1.7.2 Geometric Objects: geom_dotplot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

## The code chunk below performs the following two steps:

#scale_y_continuous() is used to turn off the y-axis, and #binwidth argument is used to change the binwidth to 2.5.

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)  
```

## 1.7.3 Geometric Objects: geom_histogram()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()       
```

## 1.7.4 Modifying a geometric object by changing geom()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")  
```

## 1.7.5 Modifying a geometric object by changing aes()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

## 1.7.6 Geometric Objects: geom-density()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()           
```

## The code chunk below plots two kernel density lines by using colour or fill arguments of aes()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

## 1.7.7 Geometric Objects: geom_boxplot

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()            
```

# Notched plot instead of boxplot

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

## 1.7.8 Geometric Objects: geom_violin

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

## 1.7.9 Geometric Objects: geom_point()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()            
```

## 1.7.10 geom objects can be combined

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)        
```

## 1.8 Essential Grammatical Elements in ggplot2: stat

#There are two ways to use these functions: #add a stat\_() function and override the default geom, or #add a geom\_() function and override the default stat.

## 1.8.1 Working with stat()

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

## 1.8.2 Working with stat - the stat_summary() method

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)               
```

## 1.8.3 Working with stat - the geom() method

```{r}
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",        
             fun="mean",           
             colour="red",          
             size=4)          
```

## 1.8.4 Adding a best fit curve on a scatterplot?

#In the code chunk below, geom_smooth() is used to plot a best fit curve on the scatterplot.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(size=0.5)
```

#The default smoothing method can be overridden as shown below.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

## 1.9 Essential Grammatical Elements in ggplot2: Facets

## 1.9.1 Working with facet_wrap()

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

## 1.9.2 facet_grid() function

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## 1.10 Essential Grammatical Elements in ggplot2: Coordinates

## 1.10.1 Working with Coordinate

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

# The code chunk below flips the horizontal bar chart into vertical bar chart by using coord_flip().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

## 1.10.2 Changing the y- and x-axis range

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5)
```

# the code chunk below fixed both the y-axis and x-axis range from 0-100.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## 1.11 Essential Grammatical Elements in ggplot2: themes

#Built-in themes include: - theme_gray() (default) - theme_bw() - theme_classic()

## 1.11.1 Working with theme

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

#A horizontal bar chart plotted using theme_classic().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

#A horizontal bar chart plotted using theme_minimal().

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```
