---
title: "Take_Home Exercise 1"
author: "Pwint Phoo Thaw"
date-modified: "last-modified"
execute:
  echo: true
  eva: true
  warning: false
  freeze: true
---

# **Ship Performance in the Gulf of Guinea**

## 1. Overview

### Setting the scene

The maritime sector is one of the most critical components of global trade, contributing significantly to economic growth and sustainability. Understanding ship performance, fuel efficiency, and operational cost factors are essential for improving decision-making and minimizing environmental impact.

An international media company that publishes weekly content on digital platforms is planning to release articles on "Ship Performance in the Gulf of Guinea".

As the role of the graphical editor of the media company, I am going to prepare data visualization for the article.

### Objective

In this exercise, Exploratory Data Analysis (EDA) methods and ggplot functions are used to explore:

-   the distribution of Ship performance in **efficiency, operational cost, speed, fuel consumption, and revenue generation**.

-   the relationship between these performances with draft, cargo weight, weather, engine type, ship type, seasonal impact and etc.

## 2. Getting Started

### Loading packages

The following R packages using the `pacman::p_load()` function are loaded.

-   **tidyverse**: Core collection of R packages designed for data science

-   **haven**: To read in data formats such as SAS and SPSS

-   **ggrepel**: to provides geoms for **ggplot2** to repel overlapping text labels

-   **ggthemes**: to use additional themes for **ggplot2**

-   **patchwork**: to prepare composite figure created using **ggplot2**

-   **ggridges**: to plot ridgeline plots

-   **ggdist**: for visualizations of distributions and uncertainty

-   **scales**: provides the internal scaling infrastructure used by **ggplot2**

```{r}
pacman::p_load(tidyverse, haven,
               ggrepel, ggthemes,
               ggridges, ggdist,
               patchwork, scales)

```

## 3. Data Wrangling

### Importing data

The dataset used in the exercise is in CSV format, retrieved from `Kaggle.com`<https://www.kaggle.com/datasets/jeleeladekunlefijabi/ship-performance-clustering-dataset>

The code chunk below import the dataset using `read_csv()` function of the haven package.

```{r}
library(readr) 
marine <- read_csv("data/Ship_Performance_Dataset.csv")
```

### Glimpse of data

Using the `glimpse()` function, we see that the dataset consists of 2,736 rows and 18 columns. It also shows the column names, column type, and the first few entries of each column.

```{r}
glimpse(marine)
```

### Checking for Duplicates

Using the `duplicated` function, we see that there are no duplicate entries in the data.

```{r}
marine[duplicated(marine),]
```

### Checking for Missing Values

Numerical Variables are selected to identify

The `colSums()` function in the **base** package is used to check for missing values in `marine`. There are no missing values in the tibble data frame.

```{r}
colSums(is.na(marine))
```

### Correlation Analysis

Numerical variables are selected to perform correlation analysis. All the variables looks fine.

```{r}
library(ggplot2)
library(dplyr)
library(reshape2)

# Select only numeric variables
numeric_vars <- marine %>% select(where(is.numeric))

# Compute correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")

# Melt correlation matrix for better visualization
cor_data <- melt(cor_matrix)

# Improved heatmap-style correlation matrix
ggplot(data = cor_data, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile(color = "white") + 
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) + 
  scale_fill_gradient2(low = "#E46726", mid = "white", high = "#6D9EC1", midpoint = 0) +
  labs(title = "Correlation Matrix of Marine Dataset",
       x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


```

### Filtering for Relevant Variables

Of 18 columns, the following 12 are preliminary selected to answer the analytical questions. The [Code Book](https://www.kaggle.com/datasets/jeleeladekunlefijabi/ship-performance-clustering-dataset/data) from Kaggle provides detailed description of the variables and I have referred to it..

+-------------------------+------------------------------------------------------------------+------------------+
| Variable Name           | Description                                                      | Type of Variable |
+=========================+==================================================================+==================+
| Ship_Type               | Type of ship                                                     | Categorical      |
|                         |                                                                  |                  |
|                         | (Fish Carrier, Container Ship, Tanker, Bulk Carrier)             |                  |
+-------------------------+------------------------------------------------------------------+------------------+
| Route_Type              | Shipping route type                                              | Categorical      |
|                         |                                                                  |                  |
|                         | (Coastal, Short-haul, Long-haul, Transoceanic)                   |                  |
+-------------------------+------------------------------------------------------------------+------------------+
| Engine_Type             | Type of engine                                                   | Categorical      |
|                         |                                                                  |                  |
|                         | (None, Diesel, Heavy Fuel Oil (HFO), Steam Turbine)              |                  |
+-------------------------+------------------------------------------------------------------+------------------+
| Weather_Condition       | Maintenance condition of the ship                                | Categorical      |
|                         |                                                                  |                  |
|                         | (None, Calm, Moderate ,Rough)                                    |                  |
+-------------------------+------------------------------------------------------------------+------------------+
| Maintenance_Status      | Prevailing weather conditions during voyages                     | Categorical      |
|                         |                                                                  |                  |
|                         | (None, Good, Fair, Critical)                                     |                  |
+-------------------------+------------------------------------------------------------------+------------------+
| Speed_Over_Ground_knots | Average speed of the ship over water                             | Numerical        |
+-------------------------+------------------------------------------------------------------+------------------+
| Engine_Power_kW         | Engine power output                                              | Numerical        |
+-------------------------+------------------------------------------------------------------+------------------+
| Distance_Traveled_nm    | Total distance traveled by the ship                              | Numerical        |
+-------------------------+------------------------------------------------------------------+------------------+
| Operational_Cost_USD    | Total operational cost per voyage                                | Numerical        |
+-------------------------+------------------------------------------------------------------+------------------+
| Revenue_per_Voyage_USD  | Revenue generated per voyage                                     | Numerical        |
+-------------------------+------------------------------------------------------------------+------------------+
| Efficiency_nm_per_kWh   | Energy efficiency calculated in nautical miles per kilowatt-hour | Numerical        |
+-------------------------+------------------------------------------------------------------+------------------+
| Average_Load_Percentage | Average load of the ship calculated in percentage                | Numerical        |
+-------------------------+------------------------------------------------------------------+------------------+

```{r}
marine_filtered <- marine %>%
  select(Ship_Type, Route_Type,Engine_Type,Weather_Condition,Maintenance_Status,Speed_Over_Ground_knots,
        Engine_Power_kW, Distance_Traveled_nm,Operational_Cost_USD,Revenue_per_Voyage_USD,Efficiency_nm_per_kWh,
        Average_Load_Percentage)

glimpse(marine_filtered)
```

## 4. Exploratory Data Analysis

### EDA 1: Distribution of Numerical Variables

```{r}
# Load necessary libraries
library(ggplot2)
library(ggridges)
library(dplyr)
library(tidyr)
library(viridis)
library(patchwork)

# Convert data to long format for visualization
marine_long <- marine %>%
  pivot_longer(
    cols = c("Speed_Over_Ground_knots", "Engine_Power_kW", "Distance_Traveled_nm", 
             "Operational_Cost_USD", "Revenue_per_Voyage_USD", "Efficiency_nm_per_kWh", 
             "Average_Load_Percentage"),
    names_to = "Variable",
    values_to = "Value"
  )

# Apply log transformation for better scaling
marine_long <- marine_long %>%
  mutate(Value = ifelse(Value > 0, log10(Value), NA))  # Log transformation

# Improved Ridge Plot (Better visualization of distributions)
marine_ridgeplot <- ggplot(marine_long, aes(x = Value, y = Variable, fill = Variable)) +
  geom_density_ridges(alpha = 0.7, quantile_lines = TRUE, 
                      quantile_fun = function(y, ...) quantile(y, probs = 0.5, na.rm = TRUE)) +
  scale_x_continuous(name = "Log-Transformed Values", labels = scales::comma) +
  scale_fill_viridis_d() +
  labs(title = "Distributions of Key Numerical Variables in Marine Dataset") +
  theme_minimal() +
  theme(legend.position = "none", axis.title.y = element_blank())

# Improved Boxplot (Detecting outliers with better scaling)
marine_boxplot <- ggplot(marine_long, aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 16, alpha = 0.8) +
  coord_flip() +  
  scale_fill_viridis_d() +
  labs(title = "Boxplots of Key Numerical Variables in Marine Dataset",
       subtitle = "Red dots indicate potential outliers",
       x = "", y = "Log-Transformed Values") +
  theme_minimal() +
  theme(legend.position = "none", axis.title.y = element_blank())

# Combine Ridge Plot and Boxplot
marine_ridgeplot / marine_boxplot


```

**Insights**

-   The data has been log-transformed, which suggests the original values had a skewed distribution (likely right-skewed).

-   Efficiency_nm_per_kWh" and "Distance_Traveled_nm" have multiple red dots, indicating several outliers.

-   "Revenue_per_Voyage_USD" and "Operational_Cost_USD" also have notable outliers, possibly reflecting extreme voyages or cost fluctuations.

### EDA 2: Exploring the distribution of categorical variables

```{r}


# Convert categorical variables into long format
marine_long <- marine %>%
  select(Ship_Type, Route_Type, Engine_Type, Weather_Condition, Maintenance_Status) %>%
  pivot_longer(cols = everything(), names_to = "Category", values_to = "Value")

# Create horizontal bar charts with different colors for categories
ggplot(marine_long, aes(y = Value, fill = Category)) +
  geom_bar(color = "black") +  # Black outline for bars
  facet_wrap(~Category, scales = "free_y") +  # Facet by category
  labs(title = "Categorical Variable Distributions", x = "Count", y = "Categories") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10))  


```

**Insights**

-   The most common engine types are **Heavy Fuel Oil (HFO) and Diesel**, suggesting that most ships operate on these fuel sources.

-   Most ships are **in Good or Fair maintenance condition**, with **fewer classified as Critical**.

-   **Long-haul and Coastal routes** also have a significant presence.

-   **Bulk Carrier, Container Ship, and Fish Carrier** are the most common ship types in the dataset.

-   Since **most ships operate in calm or moderate weather**, it could suggest that voyages are planned around favorable conditions.

### EDA 3: Exploring Relationship between Speed and Engine Power

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(patchwork)

# Create the faceted density plot (Speed Over Ground Distribution by Engine Power)
speed_densityplot <- ggplot(data = marine, aes(x = Speed_Over_Ground_knots, fill = Engine_Type)) +    
  geom_density(alpha = 0.3) + 
  facet_wrap(~ Engine_Type, scales = "free_y", nrow = 2) +  
  labs(x = "Speed Over Ground (knots)", y = "Density", 
       title = "Speed Distribution Across Engine Types",
       subtitle = "Density plot of Speed Over Ground, colored by Engine Type") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "none", 
        plot.title = element_text(face = "bold", size = 12), 
        plot.subtitle = element_text(size = 8), 
        axis.title.y = element_text(size = 8), 
        axis.title.x = element_text(size = 8))

# Create the boxplot for Speed Over Ground by Engine Type
speed_boxplot <- ggplot(marine, aes(x = Engine_Type, y = Speed_Over_Ground_knots, fill = Engine_Type)) +
  geom_boxplot(position = position_dodge(0.8)) +
  theme_minimal() +
  labs(y = "Speed Over Ground (knots)", subtitle = "Box plots of Speed Over Ground by Engine Type") +
  theme(axis.title.x = element_blank(), 
        legend.title = element_text(face = "bold", size = 8), 
        plot.subtitle = element_text(size = 8), 
        axis.title.y = element_text(size = 8)) +
  scale_fill_brewer(palette = "Set2")



# Combine the plots
speed_densityplot + speed_boxplot
```

**Insights**

-   **Diesel and Heavy Fuel Oil (HFO)** engines have unimodal distributions, meaning ships with these engines tend to operate at a consistent range of speeds.
-   **Steam Turbine:** The bimodal distribution (two peaks) suggests two common speed ranges. This could indicate different operating modes, such as low-speed cruising and high-speed transit.
-   The median speed is similar across all engine types, around 17-18 knots. This suggests that regardless of engine type, ships tend to operate within a consistent speed range.

### EDA 4: Exploring Relationship between Efficiency Across Ship Type, Engine Type and Route Type

::: panel-tabset
## Box Plot

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
# Function to create boxplots
plot_efficiency_boxplot <- function(var) {
  ggplot(marine, aes(x = .data[[var]], y = Efficiency_nm_per_kWh, fill = .data[[var]])) +
    geom_boxplot(outlier.colour = "red", outlier.shape = 16, alpha = 0.7) +
    theme_minimal() +
    labs(y = "Efficiency (nm per kWh)", x = var, 
         subtitle = paste("Boxplot of Efficiency across", var)) +
    scale_fill_brewer(palette = "Set2") +
    theme(axis.title.x = element_text(face = "bold", size = 10),
          legend.position = "none",
          plot.subtitle = element_text(size = 10),
          axis.title.y = element_text(size = 10)) +
    coord_flip()
}

# Generate boxplots
boxplot_ship <- plot_efficiency_boxplot("Ship_Type")
boxplot_engine <- plot_efficiency_boxplot("Engine_Type")
boxplot_route <- plot_efficiency_boxplot("Route_Type")

# Display all boxplots together
boxplot_ship / boxplot_engine / boxplot_route
```

## Code

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
# Function to create boxplots
plot_efficiency_boxplot <- function(var) {
  ggplot(marine, aes(x = .data[[var]], y = Efficiency_nm_per_kWh, fill = .data[[var]])) +
    geom_boxplot(outlier.colour = "red", outlier.shape = 16, alpha = 0.7) +
    theme_minimal() +
    labs(y = "Efficiency (nm per kWh)", x = var, 
         subtitle = paste("Boxplot of Efficiency across", var)) +
    scale_fill_brewer(palette = "Set2") +
    theme(axis.title.x = element_text(face = "bold", size = 10),
          legend.position = "none",
          plot.subtitle = element_text(size = 10),
          axis.title.y = element_text(size = 10)) +
    coord_flip()
}

# Generate boxplots
boxplot_ship <- plot_efficiency_boxplot("Ship_Type")
boxplot_engine <- plot_efficiency_boxplot("Engine_Type")
boxplot_route <- plot_efficiency_boxplot("Route_Type")

# Display all boxplots together
boxplot_ship / boxplot_engine / boxplot_route

```
:::

### EDA 5: Heatmap of Efficiency for different Ship_Type, Engine_Type and Route_Type

```{r}

#| fig-width: 12
#| fig-height: 10
# Convert Efficiency into categories for heatmap visualization
marine <- marine %>%
  mutate(
    Efficiency_Category = factor(case_when(
      Efficiency_nm_per_kWh >= quantile(Efficiency_nm_per_kWh, 0.75, na.rm = TRUE) ~ "High",
      Efficiency_nm_per_kWh >= quantile(Efficiency_nm_per_kWh, 0.25, na.rm = TRUE) ~ "Moderate",
      TRUE ~ "Low"
    ), levels = c("Low", "Moderate", "High"))
  )

# Calculate proportions across efficiency levels by category
efficiency_proportions <- marine %>%
  gather(key = "Efficiency_Level", value = "Value", Efficiency_Category) %>%
  count(Ship_Type, Engine_Type, Route_Type, Efficiency_Level, Value) %>%
  group_by(Ship_Type, Engine_Type, Route_Type, Efficiency_Level) %>%
  mutate(Proportion = n / sum(n),
         Value = factor(Value, levels = c("Low", "Moderate", "High"))) %>%
  ungroup()

# Function to create heatmaps
plot_efficiency_heatmap <- function(var) {
  ggplot(efficiency_proportions, aes(x = .data[[var]], y = Value, fill = Proportion)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "steelblue") +
    facet_wrap(~ Efficiency_Level, scales = "free_x") +
    labs(x = var,
         y = "Efficiency Level",
         fill = "Proportion",
         subtitle = paste("Heatmap of Efficiency in", var)) +
    theme_minimal() +
    theme(panel.background = element_rect(fill = "white", colour = NA),
          axis.text.x = element_text(size = 8, angle = 45), 
          axis.text.y = element_text(size = 8),
          axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8), 
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 8),
          plot.subtitle = element_text(size = 8))
}

# Generate heatmaps
efficiency_heatmap_ship <- plot_efficiency_heatmap("Ship_Type")
efficiency_heatmap_engine <- plot_efficiency_heatmap("Engine_Type")
efficiency_heatmap_route <- plot_efficiency_heatmap("Route_Type")

# Display all heatmaps together
efficiency_heatmap_ship + efficiency_heatmap_engine + efficiency_heatmap_route

```

**Insights**

-   Moderate efficiency dominates across all categories, suggesting that most ships operate within standard efficiency ranges.

-   Bulk Carriers, Container Ships, Diesel, and HFO engines tend to be more efficient.

-   Fish Carriers, Steam Turbines, and Coastal Routes show greater efficiency variability, suggesting they are more sensitive to operational factors.

### EDA 6: Exploring Financial Performance

```{r}
library(ggplot2)
library(dplyr)
library(ggridges)
library(patchwork)

# Compute Profitability
marine <- marine %>%
  mutate(Profitability_USD = Revenue_per_Voyage_USD - Operational_Cost_USD)

# Ridgeline Plot of Profitability by Ship Type
profitability_ridgeline <- ggplot(marine, aes(x = Profitability_USD, y = Ship_Type, fill = 0.5 - abs(0.5 - stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient", calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability", direction = -1) +
  labs(x = "Profitability (USD)", y = "Ship Type", 
       title = "Profitability Distribution by Ship Type",
       subtitle = "Ridgeline plot comparing density distributions") +
  theme_minimal() +
  theme(legend.position = "none", 
        plot.title = element_text(face = "bold", size = 12), 
        plot.subtitle = element_text(size = 10),
        axis.title.y = element_text(size = 10), 
        axis.title.x = element_text(size = 10))

# Display Ridgeline Plot
print(profitability_ridgeline)
```

**Insights**

-   **Most ship types operate near break-even profitability**, with some variation in earnings.\

-   **Bulk Carriers and Container Ships** show more stable profitability trends.\

-   **Tanker ships have a wider range of profitability,** potentially reflecting volatile fuel costs and market conditions.

### **EDA 7:** Treemap of Revenue Contribution by Ship Type

```{r}
# Load required libraries
library(treemap)
library(dplyr)



# Summarize data: Total revenue & average load percentage per Ship Type & Route Type
treemap_data <- marine %>%
  group_by(Ship_Type, Route_Type) %>%
  summarise(
    Total_Revenue = sum(Revenue_per_Voyage_USD, na.rm = TRUE),
    Avg_Load = mean(Average_Load_Percentage, na.rm = TRUE)
  ) %>%
  ungroup()

# Generate treemap with color based on Ship_Type
tm <- treemap(treemap_data,
        index = c("Ship_Type", "Route_Type"),  # Hierarchical structure
        vSize = "Total_Revenue",  # Size based on total revenue
        vColor = "Ship_Type",  # Color by ship type (categorical)
        type = "categorical",  # Use categorical color scheme
        palette = "Set3",  # More distinct colors for ship types
        title = "Revenue Contribution by Ship Type and Route",
        title.legend = "Ship Type"
)


```

-   **Insights**

-   **Bulk Carriers (green) and Container Ships (yellow) have the largest sections,** indicating they generate the most revenue.

-   Within these ship types, **Long-haul and Coastal routes contribute significantly.**

-   **Transoceanic routes for Bulk Carriers** are also substantial.

## 5. Summary and Conclusion

From the analysis of ship performance, efficiency, profitability, and revenue distribution, the following key **implications** emerge:

-   **Most voyages occur under stable weather conditions**, and **route type significantly impacts operational choices**.

-   The **choice of engine type influences speed consistency**, with **Steam Turbines being more variable** and **Diesel/HFO engines maintaining steady performance**.

-   Profitability is **highly dependent on ship type and operational costs**. **Container Ships and Bulk Carriers** show **consistent profits**, while Tankers **face more financial risk due to market volatility**.

-   Shipping companies should **prioritize investments in Bulk Carriers and Container Ships** while optimizing **Long-haul and Coastal routes**. **Short-haul voyages might need alternative revenue models** or pricing strategies.
