---
title: "Hands-on_Ex08: Visualising and Analysing Geographic Data"
format: html
date-modified: "last-modified"
execute:
  echo: true
  eva: true
  warning: false
  freeze: true
---

# 8 Choropleth Mapping with R

## 8.1 Overview

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colors. For example, a social scientist may need to use a choropleth map to portray the spatial distribution of aged population of Singapore by Master Plan 2014 Subzone Boundary.

## 8.2 Getting Started

In this hands-on exercise, the key R package use is [**tmap**](https://cran.r-project.org/web/packages/tmap/) package in R. Beside **tmap** package, four other R packages will be used. They are:

-   [**readr**](https://readr.tidyverse.org/) for importing delimited text file,

-   [**tidyr**](https://tidyr.tidyverse.org/) for tidying data,

-   [**dplyr**](https://dplyr.tidyverse.org/) for wrangling data and

-   [**sf**](https://cran.r-project.org/web/packages/sf/index.html) for handling geospatial data.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## **8.3 Importing Data into R**

Two data set will be used to create the choropleth map. They are:

-   Master Plan 2014 Subzone Boundary (Web) (i.e. `MP14_SUBZONE_WEB_PL`) in ESRI shapefile format. It can be downloaded at [data.gov.sg](https://data.gov.sg/) This is a geospatial data. It consists of the geographical boundary of Singapore at the planning subzone level. The data is based on URA Master Plan 2014.

-   Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 in csv format (i.e. `respopagesextod2011to2020.csv`). This is an aspatial data fie. It can be downloaded at [Department of Statistics, Singapore](https://www.singstat.gov.sg/) Although it does not contain any coordinates values, but it’s PA and SZ fields can be used as unique identifiers to geocode to `MP14_SUBZONE_WEB_PL` shapefile.

-   The code chunk below uses the *st_read()* function of **sf** package to import `MP14_SUBZONE_WEB_PL` shapefile into R as a simple feature data frame called `mpsz`.

```{r}
mpsz <- st_read(dsn = "data/geospatial",                  
                layer = "MP14_SUBZONE_WEB_PL")
```

You can examine the content of `mpsz` by using the code chunk below.

```{r}
mpsz
```

### **8.3.3 Importing Attribute Data into R**

Next, we will import *respopagsex2011to2020.csv* file into RStudio and save the file into an R dataframe called *popagsex*.

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

### **8.3.4 Data Preparation**

Before a thematic map can be prepared, you are required to prepare a data table with year 2020 values. The data table should include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

#### 8.3.4.1 Data wrangling

The following data wrangling and transformation functions will be used:

-   *pivot_wider()* of **tidyr** package, and

-   *mutate()*, *filter()*, *group_by()* and *select()* of **dplyr** package

```{r}
    popdata2020 <- popdata %>%
      filter(Time == 2020) %>%
      group_by(PA, SZ, AG) %>%
      summarise(`POP` = sum(`Pop`)) %>%
      ungroup() %>%
      pivot_wider(names_from=AG, 
                  values_from=POP) %>%
      mutate(YOUNG = rowSums(.[3:6])
             +rowSums(.[12])) %>%
    mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
    rowSums(.[13:15]))%>%
    mutate(`AGED`=rowSums(.[16:21])) %>%
    mutate(`TOTAL`=rowSums(.[3:21])) %>%  
    mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
    /`ECONOMY ACTIVE`) %>%
      select(`PA`, `SZ`, `YOUNG`, 
           `ECONOMY ACTIVE`, `AGED`, 
           `TOTAL`, `DEPENDENCY`)
```

```         
#### 8.3.4.2 Joining the attribute data and geospatial data

Before we can perform the georelational join, one extra step is required to convert the values in PA and SZ fields to uppercase. This is because the values of PA and SZ fields are made up of upper- and lowercase. 
```

```{r}
    popdata2020 <- popdata2020 %>%
      mutate(across(c(PA, SZ), toupper)) %>%  # Convert PA and SZ to uppercase
      filter(`ECONOMY ACTIVE` > 0)  # Keep only rows where ECONOMY ACTIVE > 0


```

Next, *left_join()* of **dplyr** is used to join the geographical data and attribute table using planning subzone name e.g. *SUBZONE_N* and *SZ* as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

## **8.4 Choropleth Mapping Geospatial Data Using *tmap***

### **8.4.1 Plotting a choropleth map quickly by using *qtm()***

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

Things to learn from the code chunk above:

-   *tmap_mode()* with “plot” option is used to produce a static map. For interactive mode, “view” option should be used.

-   *fill* argument is used to map the attribute (i.e. DEPENDENCY)

### **8.4.2 Creating a choropleth map by using *tmap*’s elements**

To draw a high quality cartographic choropleth map as shown in the figure below, **tmap**’s drawing elements should be used.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

#### 8.4.2.1 Drawing a base map

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

#### 8.4.2.2 Drawing a choropleth map using *tm_polygons()*

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

#### 8.4.2.3 Drawing a choropleth map using *tm_fill()* and \*tm_border()\*\*

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

Notice that the planning subzones are shared according to the respective dependecy values

To add the boundary of the planning subzones, tm_borders will be used as shown in the code chunk below.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") + tm_borders(lwd = 0.1,  alpha = 1,lty ='solid')
  
```

### **8.4.3 Data classification methods of tmap**

#### 8.4.3.1 Plotting choropleth maps with built-in classification methods

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

In the code chunk below, *equal* data classification method is used.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 2,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 10,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 20,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

### **Key Takeaways**

-   **Higher `n` (more bins)** → More detailed visualization, but might introduce unnecessary complexity.

-   **Lower `n` (fewer bins)** → Simpler visualization, but may hide finer variations in data.

#### 8.4.3.2 Plotting choropleth map with custome break

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

With reference to the results above, we set break point at 0.60, 0.70, 0.80, and 0.90. In addition, we also need to include a minimum and maximum, which we set at 0 and 100. Our *breaks* vector is thus c(0, 0.60, 0.70, 0.80, 0.90, 1.00)

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

### **8.4.4 Colour Scheme**

**tmap** supports colour ramps either defined by the user or a set of predefined colour ramps from the **RColorBrewer** package.

#### 8.4.4.1 Using ColourBrewer palette

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "Blues") +
  tm_borders(alpha = 0.5)
```

To reverse the colour shading, add a “-” prefix.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5)
```

### **8.4.5 Map Layouts**

Map layout refers to the combination of all map elements into a cohensive map. Map elements include among others the objects to be mapped, the title, the scale bar, the compass, margins and aspects ratios.

#### 8.4.5.1 Map Legend

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "jenks", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

#### 8.4.5.2 Map style

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

#### 8.4.5.3 Cartographic Furniture

Beside map style, **tmap** also also provides arguments to draw other map furniture such as compass, scale bar and grid lines.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio \nby planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar(breaks = c(0, 5, 10)) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

### **8.4.6 Drawing Small Multiple Choropleth Maps**

**Small multiple maps**, also referred to as **facet maps**, are composed of many maps arrange side-by-side, and sometimes stacked vertically. Small multiple maps enable the visualisation of how spatial relationships change with respect to another variable, such as time.

#### 8.4.6.1 By assigning multiple values to at least one of the aesthetic arguments

In this example, small multiple choropleth maps are created by defining ***ncols*** in **tm_fill()**

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

In this example, small multiple choropleth maps are created by assigning multiple values to at least one of the aesthetic arguments

```{r}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"), 
          palette = list("Blues","-Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

#### 8.4.6.3 By creating multiple stand-alone maps with *tmap_arrange()*

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### **8.4.7 Mappping Spatial Object Meeting a Selection Criterion**

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)

```

# **Part 2 of 8 Visualising Geospatial Point Data**

## **8 Overview**

Proportional symbol maps (also known as graduate symbol maps) are a class of maps that use the visual variable of size to represent differences in the magnitude of a discrete, abruptly changing phenomenon, e.g. counts of people.

## **8.1 The data**

The data set use for this hands-on exercise is called *SGPools_svy21*. The data is in csv file format.

Figure below shows the first 15 records of SGPools_svy21.csv. It consists of seven columns. The XCOORD and YCOORD columns are the x-coordinates and y-coordinates of SingPools outlets and branches. They are in [Singapore SVY21 Projected Coordinates System](https://www.sla.gov.sg/sirent/CoordinateSystems.aspx).

### **8.1.1 Data Import and Preparation**

```{r}
sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
```

```{r}
list(sgpools) 
```

### **8.1.2 Creating a sf data frame from an aspatial data frame**

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

```{r}
list(sgpools_sf)
```

## 8.2 **Drawing Proportional Symbol Map**

```{r}
tmap_mode("view")
```

### **8.2.1 all started with an interactive point symbol map**

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "red",
           size = 1,
           border.col = "black",
           border.lwd = 1)
```

## **8.3 Lets make it proportional**

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "red",
           size = "Gp1Gp2 Winnings",
           border.col = "black",
           border.lwd = 1)
```

## **8.4 Lets give it a different colour**

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1)
```

## **8.5 I have a twin brothers :)**

An impressive and little-know feature of **tmap**’s view mode is that it also works with faceted plots. The argument *sync* in *tm_facets()* can be used in this case to produce multiple maps with synchronised zoom and pan settings.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

```{r}
tmap_mode("plot")
```

# **Part 3 of 8  Analytical Mapping**

## **8.1 Overview**

By the end of this in-class exercise, you will be able to use appropriate functions of tmap and tidyverse to perform the following tasks:

-   Importing geospatial data in rds format into R environment.

-   Creating cartographic quality choropleth maps by using appropriate tmap functions.

-   Creating rate map

-   Creating percentile map

-   Creating boxmap

## **8.2 Getting Started**

### **8.2.2 Importing data**

```{r}
NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

## **8.3 Basic Choropleth Mapping**

### **8.3.1 Visualising distribution of non-functional water point**

```{r}
p1 <- tm_shape(NGA_wp) +
  tm_fill("wp_functional",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of functional water point by LGAs",
            legend.outside = FALSE)
```

```{r}
p2 <- tm_shape(NGA_wp) +
  tm_fill("total_wp",
          n = 10,
          style = "equal",
          palette = "Blues") +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Distribution of total  water point by LGAs",
            legend.outside = FALSE)
```

```{r}
tmap_arrange(p2, p1, nrow = 1)
```

## **8.4 Choropleth Map for Rates**

### **8.4.1 Deriving Proportion of Functional Water Points and Non-Functional Water Points**

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

### **8.4.2 Plotting map of rate**

```{r}
tm_shape(NGA_wp) +
  tm_fill("pct_functional",
          n = 10,
          style = "equal",
          palette = "Blues",
          legend.hist = TRUE) +
  tm_borders(lwd = 0.1,
             alpha = 1) +
  tm_layout(main.title = "Rate map of functional water point by LGAs",
            legend.outside = TRUE)
```

## **8.5 Extreme Value Maps**

### **8.5.1 Percentile Map**

The percentile map is a special type of quantile map with six specific categories: 0-1%,1-10%, 10-50%,50-90%,90-99%, and 99-100%. The corresponding breakpoints can be derived by means of the base R quantile command, passing an explicit vector of cumulative probabilities as c(0,.01,.1,.5,.9,.99,1).

#### 8.5.1.1 Data Preparation

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

Step 2: Creating customised classification and extracting values

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

#### 8.5.1.3 Creating the get.var function

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}

```

#### 8.5.1.4 A percentile mapping function

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
  tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,
             title=legtitle,
             breaks=bperc,
             palette="Blues",
          labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("right","bottom"))
}
```

#### 8.5.1.5 Test drive the percentile mapping function

```{r}
percentmap("total_wp", NGA_wp)
```

### **8.5.2 Box map**

```{r}
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

-   Displaying summary statistics on a choropleth map by using the basic principles of boxplot.

-   To create a box map, a custom breaks specification will be used. However, there is a complication. The break points for the box map vary depending on whether lower or upper outliers are present.

#### 8.5.2.1 Creating the boxbreaks function

```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

#### 8.5.2.2 Creating the get.var function

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

#### 8.5.2.3 Test drive the newly created function

```{r}
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

#### 8.5.2.4 Boxmap function

```{r}
boxmap <- function(vnam, df, 
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
          labels = c("lower outlier", 
                     "< 25%", 
                     "25% - 50%", 
                     "50% - 75%",
                     "> 75%", 
                     "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle, 
            title.position = c("left",
                               "top"))
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
