---
title: "In-class_Ex07: Visualising, Analysing and Forecasting Time-series Data: tidyverts methods"
format: html
date-modified: "last-modified"
execute:
  echo: true
  eva: true
  warning: false
  freeze: true
---

## **7.2 Getting Started**

```{r}
pacman::p_load(tidyverse, tsibble, feasts, fable, seasonal)
```

### **7.2.1 Importing the data**

```{r}
ts_data <- read_csv(
  "visitor_arrivals_by_air.csv")
```

```{r}
ts_data$`Month-Year` <- dmy(
  ts_data$`Month-Year`)
```

```{r}
ts_data
```

### **7.2.3 Conventional base `ts` object versus `tibble` object**

```{r}
ts_data_ts <- ts(ts_data)       
head(ts_data_ts)
```

### **7.2.4 Converting `tibble` object to `tsibble` object**

```{r}
ts_tsibble <- ts_data %>%
  mutate(Month = yearmonth(`Month-Year`)) %>%
  as_tsibble(index = `Month`)
```

## **7.3 Visualising Time-series Data**

```{r}
ts_longer <- ts_data %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```

### **7.3.1 Visualising single time-series: ggplot2 methods**

```{r}
ts_longer %>%
  filter(Country == "Vietnam") %>%
  ggplot(aes(x = `Month-Year`, 
             y = Arrivals))+
  geom_line(size = 0.5)
```

```{r}
ts_longer %>%
  filter(Country == "Malaysia") %>%
  ggplot(aes(x = `Month-Year`, 
             y = Arrivals))+
  geom_line(size = 0.5)
```

```{r}
ts_longer %>%
  filter(Country == "Myanmar") %>%
  ggplot(aes(x = `Month-Year`, 
             y = Arrivals))+
  geom_line(size = 0.5)
```

### **9.3.2 Plotting multiple time-series data with ggplot2 methods**

very hard to see

```{r}
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals,
           color = Country))+
  geom_line(size = 0.5) +
  theme(legend.position = "bottom", 
        legend.box.spacing = unit(0.5, "cm"))
```

In order to provide effective comparison, [`facet_wrap()`](https://ggplot2.tidyverse.org/reference/facet_wrap.html) of **ggplot2** package is used to create small multiple line graph also known as trellis plot.

```{r}
#| echo: false
#| fig-height: 12
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals))+
  geom_line(size = 0.5) +
  facet_wrap(~ Country,
             ncol = 3,  #after defining 3 coloumn, next one will go down to row
             scales = "free_y") +
  theme_bw()
```
